Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2013-01-16 many.tex
Minor edits to IV.

To begin with, we recall the well-known fact that the Bayesian
estimator is unbiased, in the following sense.  For simplicity, assume
that the model consists of a single rate parameter $R$, with prior
distribution $p(R)$.  Consider an ensemble of data sets whose
distribution is consistent with that prior; i.e., such that $p(d)$ is
given by
%Construct and ensemble of synthetic data sets consistent with $p(R)$; i.e., draw data from the %distribution
\begin{equation}
 p(d) = \int{ p(d|R)\,p(R)\, dR}\, .  
\end{equation}
For each data set in the ensemble, compute the Bayesian estimator
$R_{B} = \int{ R p(R|d) dR}$.  Then it is immediate that
\begin{equation}
\int{R_{B}(d)\, p(d)\,d(d) }= \int{ R\, p(R) \,dR},
\end{equation}
i.e.\ the data-weighted average of the Bayesian estimate $R_{B}$
equals the prior-weighted average $R$.  Therefore all threshold values
will yield, on average, the same estimate of the rate.  However this
equality of averages does {\it not} imply that all threshold values
yield the same information. In general, as the threshold is lowered to
include more events, the error bar on the estimate shrinks.  In this
subsection we give quantitative illustrations of how the error bar
shrinks when the threshold is lowered.

Consider the following model problem.  Let $p(x) = b(x) + f(x) = R_b
\hat{b}(x) + R_f \hat{f}(x)$ be the rate density of events (of both
foreground and background type) per unit loudness.  Here we will
assume that the background is normally-distributed in loudness, so
that $b$ has the form
\begin{equation}
b(x) = \Lambda_b \exp \left(- \frac{x^2}{2} \right).
\end{equation}
We find it useful to define $x_1$ as the loudness such that a data set
will have on average a single noise event louder than $x_1$; i.e.,
such that 
\begin{equation}
 \int_{x_1}^{\infty} b(x) dx = R_b - B\left(x_1\right) = 1.
\end{equation}
This condition fixes 
\begin{equation}
  \Lambda_b = \left[ \sqrt{\frac{\pi}{2}} \erfc\left(
    \frac{x_1}{\sqrt{2}} \right) \right]^{-1},
\end{equation}
while $R_b$ will depend on the threshold, $x_\mathrm{th}$, as 
\begin{equation}
  R_b = \frac{\erfc\left( \frac{x_\mathrm{th}}{\sqrt{2}}
    \right)}{\erfc\left( \frac{x_1}{\sqrt{2}} \right)}.
\end{equation}

Let the foreground distribution follow a power law in loudness (this
is, for example, the distribution of SNR for gravitational wave events
from uniformly-distributed sources in a single detector)
\begin{equation}
\label{eq:ps}
f(x) = 3 \Lambda_f \frac{x_1^3}{x^4},
\end{equation}
where $\Lambda_f = R_f - F\left(x_1\right)$ is the mean number of
foreground events with $x > x_1$.  The overall foreground rate is
given by 
\begin{equation}
  \label{eq:foreground-from-lambda}
  R_f = \Lambda_f \frac{x_1^3}{x_\mathrm{th}^3}.
\end{equation}

We can write the full $p(x)$ as
\begin{equation}
\label{eq:ptot} 
p(x) = \left[\sqrt{\frac{\pi}{2}}
  \erfc\left(\frac{x_1}{\sqrt{2}}\right)\right]^{-1}\exp\left[-\frac{x^2}{2}\right]
+ 3 \Lambda_f \frac{x_1^3}{x^4}
\end{equation}

For any pair $(x_1, \Lambda_f)$, it is straightforward to construct
random event lists drawn from the corresponding $p(x)$, and
straightforward to apply a threshold by ``throwing away'' all events
with $x$ less than the threshold value $x_\mathrm{th}$.  If $\Lambda_f
\gg 1$, then we are in the foreground-dominated regime at $x = x_1$,
if $\Lambda_f \ll 1$ we are in the background-dominated regime, and if
$\Lambda_f \sim 1$ the foreground and background counts above $x_1$
are about equal.  For any thresholded event list, we use
Eq.~(\ref{eq:rate-shape-posterior}) to construct the probability
density $p(R_f|d)$.  For that event list, we define the foreground
rate uncertainty, $\Delta R_f$, by
\begin{equation}
\label{eq:delta-R-definition}
(\Delta R_f)^2 \equiv \int{(R_f - R_f^\mathrm{true})^2 \, p(R_f|d)\, dR_f },
\end{equation}
where $R_f^\mathrm{true}$ is given by
Eq.~\eqref{eq:foreground-from-lambda}.

\begin{figure}
  \includegraphics[width=\columnwidth]{threshold.pdf}
  \caption{\label{fig:threshold-rate-dependence} The mean foreground
    rate uncertainty, Eq.~\eqref{eq:delta-R-definition}, as a function
    of threshold for data sets with $\Lambda_f = 100$ (solid line),
    $\Lambda_f = 10$ (dashed line), and $\Lambda_f = 1$ (dash-dotted
    line).  Recall that $\Lambda_f$ is the mean number of foreground
    events above $x_\mathrm{th} = 8$.  The total background rate,
    $R_b\left( x_\mathrm{th} \right)$, is shown by the dotted line; we
    fix $R_b \left( x_\mathrm{th} = x_1 = 8\right) = 1$, so on average
    there is one background event above $x = 8$.  For $x_\mathrm{th}
    \gtrsim x_1$, increasing the threshold tends to increase the
    foreground rate uncertainty because the rate is
    foreground-dominated and fewer events are included in the data
    set.  For $x_\mathrm{th} \lesssim x_1$, the background rate
    dominates at small loudness, and the foreground rate uncertainty
    asymptotes to the counting error on the events that stand out from
    the background, $\Delta R_f / R_f \simeq 1/\sqrt{\Lambda_f}$.
    Note that this uncertainty applies even when the total number of
    background events is orders of magnitude larger than the number of
    foreground events.  When a threshold must be chosen, it is
    safest---in the sense of producing the minimal foreground rate
    uncertainty---to choose the threshold \emph{well into the
      background-dominated loudness regime}; the extra background
    events in the data set do not affect the estimate of the
    foreground rate, and, when the background distribution is
    parameterized, can help to better determine these parameters (see
    \S~\ref{sec:gw-overlapping-template}). }
\end{figure}

Figure \ref{fig:threshold-rate-dependence} illustrates how the mean
fractional foreground uncertainty, $\left\langle \Delta R_f
\right\rangle / R_f$, varies with the threshold value $x_\mathrm{th}$
for the foreground-dominated and comparable-rate regime.  In all cases
we assumed that $x_1 = 8$.  For large thresholds, where $R_b \ll 1$,
increasing the threshold tends to increase the fractional uncertainty
on the foreground rate, since fewer foreground events are included in
the sample.  However, as the threshold passes into the
background-dominated regime, the uncertainty in the foreground rate
asymptotes to
\begin{equation}
  \frac{\Delta R_f}{R_f} \simeq \frac{1}{\sqrt{\Lambda_f}},
\end{equation}
which is the usual Poisson counting uncertainty on the events that
stand out from the background (those with $x \gtrsim x_1$).  Note that
this uncertainty applies even when the total number of background
events is orders of magnitude larger than the number of foreground
events.  When a threshold must be chosen, it is safest---in the sense
of producing the minimal foreground rate uncertainty---to choose the
threshold \emph{well into the background-dominated loudness regime};
the extra background events in the data set do not affect the estimate
of the foreground rate, and, when the background distribution is
parameterized, can help to better determine these parameters (see
\S~\ref{sec:gw-overlapping-template}).

Though we have only illustrated the behavior of the rate estimate
quantitatively for this specific example of foreground and background
rates, the conclusions hold in general.  Consider the Fisher
information matrix for the posterior distribution in
Eq.~\eqref{eq:rate-shape-posterior}.  For a model with parameters
$\mathset{\theta_i}$, the Fisher information matrix has components
\begin{equation}
  F_{ij} \equiv \left\langle \frac{\partial \log p\left( \theta | d
    \right)}{\partial \theta_i} \frac{\partial \log p\left( \theta | d
    \right)}{\partial \theta_j} \right\rangle,
\end{equation}
where the average is taken over the data distribution at fixed
$\theta$, $p(d | \theta)$.  The components of the Fisher information
matrix describe the maximum amount of information about the
corresponding parameters available in a given data set; the inverse of
the Fisher information matrix gives the Cramer-Rao bound on the
covariance matrix of unbiased estimators of $\theta$.  For the
likelihood that enters Eq.~\eqref{eq:rate-shape-posterior}, the Fisher
information matrix is 
\begin{multline}
  \mathbf{F} = \left( R_f + R_b \right) \\ \times \begin{pmatrix}
    \left \langle \left(\frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}}\right)^2 \right
    \rangle & \left \langle \frac{\hat{f} \hat{b}}{\left( R_f \hat{f}
      + R_b \hat{b} \right)^2} \right\rangle \\
    \left \langle \frac{\hat{f} \hat{b}}{\left( R_f \hat{f}
      + R_b \hat{b} \right)^2} \right\rangle & \left \langle
    \left(\frac{\hat{b}}{R_f \hat{f} + R_b \hat{b}}\right)^2 \right 
    \rangle
  \end{pmatrix},
\end{multline}
where the expectation values are taken over the distributions
$\hat{f}$ and $\hat{b}$ (i.e.\ they are expectations for one event
from the combined rate distribution).  If the cross-terms are small,
then the Cramer-Rao bound on the uncertainty of $R_f$ will be given by
\begin{equation}
  \sigma_{R_f} \simeq \frac{1}{\sqrt{R_f + R_b}} \left[ \left\langle \left(
    \frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}} \right)^2 \right\rangle\right]^{-1/2}
\end{equation}
Extending a threshold into regions where the factor 
\begin{equation}
  \left( \frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}} \right)^2
\end{equation}
becomes small---that is, into background-dominated
regions---contributes little to reducing the overall uncertainty in the
foreground rate.  This is consistent with the behavior of the specific
example in Figure \ref{fig:threshold-rate-dependence}.

@@

To explain this, we will use the same model as described in the
previous subsection, and we will begin with a very specific example.
Let $\Lambda_f = 1$, meaning that the expected number of actual events
with $x > x_1$ is one.  Then there is a $1/64$ chance ($1-\hat{F}(x_{LE}) \approx
1.6\%$) that the loudest event will have $x_{LE} > 4 x_1$.  Consider
this case, and let us also assume that there are no events (noise or
actual) with $x_1 < x < x_{LE}$.

@@

Now, if the true rate really were $\Lambda_f = 64$, then the expected
number of events with $x > x_1$ would be $64$.  So in this case, the
loudest event estimate ignores the fact that there are $\sim 56 -72$
``missing'' events.  However a Bayesian estimate with $x_\mathrm{th}$
set to $x_1$ incorporates this information quite naturally, and so
(correctly) yields an estimated $\Lambda_f$ of order one.

>>>>

To begin with, we recall the well-known fact that the Bayesian
estimator is unbiased, in the following sense.  For simplicity, assume
that the model consists of a single rate parameter $R$, with prior
distribution $p(R)$.  Consider an ensemble of data sets whose
distribution is consistent with that prior; i.e., such that $p(d)$ is
given by
%Construct and ensemble of synthetic data sets consistent with $p(R)$; i.e., draw data from the %distribution
\begin{equation}
 p(d) = \int{ p(d|R)\,p(R)\, dR}\, .  
\end{equation}
For each data set in the ensemble, compute the Bayesian estimator 
for the mean of the posterior $R_{B} = \int{ R p(R|d) dR}$.  Then it is immediate that
\begin{equation}
\int{R_{B}(d)\, p(d)\,d(d) }= \int{ R\, p(R) \,dR},
\end{equation}
i.e.\ the data-weighted average of the Bayesian estimate $R_{B}$
equals the prior-weighted average $R$.  Therefore all threshold values
will yield, on average, the same estimate of the rate.  However this
equality of averages does {\it not} imply that all threshold values
yield the same information. In general, as the threshold is lowered to
include more events, the error bar on the estimate shrinks.  In this
subsection we give quantitative illustrations of how the error bar
shrinks when the threshold is lowered.

Consider the following model problem.  Let $p(x) = b(x) + f(x) = R_b
\hat{b}(x) + R_f \hat{f}(x)$ be the rate density of events (of both
foreground and background type) per unit loudness.  Here we will
assume that the background is normally-distributed in loudness, so
that $b$ has the form
\begin{equation}
b(x) = \Gamma_b \exp \left(- \frac{x^2}{2} \right).
\end{equation}
We find it useful to define $x_1$ as the loudness such that a data set
will have on average a single noise event louder than $x_1$; i.e.,
such that 
\begin{equation}
 \int_{x_1}^{\infty} b(x) dx = R_b - B\left(x_1\right) = 1.
\end{equation}
This condition fixes 
\begin{equation}
  \Gamma_b = \left[ \sqrt{\frac{\pi}{2}} \erfc\left(
    \frac{x_1}{\sqrt{2}} \right) \right]^{-1},
\end{equation}
while $R_b$ will depend on the threshold, $x_\mathrm{th}$, as 
\begin{equation}
  R_b = \frac{\erfc\left( \frac{x_\mathrm{th}}{\sqrt{2}}
    \right)}{\erfc\left( \frac{x_1}{\sqrt{2}} \right)}.
\end{equation}

Let the foreground distribution follow a power law in loudness (this
is, for example, the distribution of SNR for gravitational wave events
from uniformly-distributed sources in a single detector)
\begin{equation}
\label{eq:ps}
f(x) = 3 \Gamma_f \frac{x_1^3}{x^4},
\end{equation}
where $\Gamma_f = R_f - F\left(x_1\right)$ is the mean number of
foreground events with $x > x_1$.  The overall foreground rate is
given by 
\begin{equation}
  \label{eq:foreground-from-lambda}
  R_f = \Gamma_f \frac{x_1^3}{x_\mathrm{th}^3}.
\end{equation}

We can write the full $p(x)$ as
\begin{equation}
\label{eq:ptot} 
p(x) = \left[\sqrt{\frac{\pi}{2}}
  \erfc\left(\frac{x_1}{\sqrt{2}}\right)\right]^{-1}\exp\left[-\frac{x^2}{2}\right]
+ 3 \Gamma_f \frac{x_1^3}{x^4}
\end{equation}

For any pair $(x_1, \Gamma_f)$, it is straightforward to construct
random event lists drawn from the corresponding $p(x)$, and
straightforward to apply a threshold by ``throwing away'' all events
with $x$ less than the threshold value $x_\mathrm{th}$.  If $\Gamma_f
\gg 1$, then we are in the foreground-dominated regime at $x = x_1$,
if $\Gamma_f \ll 1$ we are in the background-dominated regime, and if
$\Gamma_f \sim 1$ the foreground and background counts above $x_1$
are about equal.  For any thresholded event list, we use
Eq.~(\ref{eq:rate-shape-posterior}) to construct the probability
density $p(R_f|d)$.  For that event list, we define the foreground
rate uncertainty, $\Delta R_f$, by
\begin{equation}
\label{eq:delta-R-definition}
(\Delta R_f)^2 \equiv \int{(R_f - R_f^\mathrm{true})^2 \, p(R_f|d)\, dR_f },
\end{equation}
where $R_f^\mathrm{true}$ is given by
Eq.~\eqref{eq:foreground-from-lambda}.

\begin{figure}
  \includegraphics[width=\columnwidth]{threshold.pdf}
  \caption{\label{fig:threshold-rate-dependence} The mean foreground
    rate uncertainty, Eq.~\eqref{eq:delta-R-definition}, as a function
    of threshold for data sets with $\Gamma_f = 100$ (solid line),
    $\Gamma_f = 10$ (dashed line), and $\Gamma_f = 1$ (dash-dotted
    line).  Recall that $\Gamma_f$ is the mean number of foreground
    events above $x_\mathrm{th}=x_1= 8$.  The total background rate,
    $R_b\left( x_\mathrm{th} \right)$, is shown by the dotted line; we
    fix $R_b \left( x_\mathrm{th} = x_1 = 8\right) = 1$, so on average
    there is one background event above $x = 8$.  For $x_\mathrm{th}
    \gtrsim x_1$, increasing the threshold tends to increase the
    foreground rate uncertainty because the rate is
    foreground-dominated and fewer events are included in the data
    set.  For $x_\mathrm{th} \lesssim x_1$, the background rate
    dominates at small loudness, and the foreground rate uncertainty
    asymptotes to the counting error on the events that stand out from
    the background, $\Delta R_f / R_f \simeq 1/\sqrt{\Gamma_f}$.}
\end{figure}

Figure \ref{fig:threshold-rate-dependence} illustrates how the mean
fractional foreground uncertainty, $\left\langle \Delta R_f
\right\rangle / R_f$, varies with the threshold value $x_\mathrm{th}$
for the foreground-dominated and comparable-rate regime.  In all cases
we assumed that $x_1 = 8$.  For large thresholds, where $R_b \ll 1$,
increasing the threshold tends to increase the fractional uncertainty
on the foreground rate, since fewer foreground events are included in
the sample.  However, as the threshold passes into the
background-dominated regime, the uncertainty in the foreground rate
asymptotes to
\begin{equation}
  \frac{\Delta R_f}{R_f} \simeq \frac{1}{\sqrt{\Gamma_f}},
\end{equation}
which is the usual Poisson counting uncertainty on the events that
stand out from the background (those with $x \gtrsim x_1$).  Note that
this uncertainty applies even when the total number of background
events is orders of magnitude larger than the number of foreground
events.  When a threshold must be chosen, it is safest---in the sense
of producing the minimal foreground rate uncertainty---to choose the
threshold \emph{well into the background-dominated loudness regime};
the extra background events in the data set do not affect the estimate
of the foreground rate, and, when the background distribution is
parameterized, can help to better determine these parameters (see
\S~\ref{sec:gw-overlapping-template}).

Though we have only illustrated the behavior of the rate estimate
quantitatively for this specific example of foreground and background
rates, the conclusions hold in general.  Consider the Fisher
information matrix for the posterior distribution in
Eq.~\eqref{eq:rate-shape-posterior}.  For a model with parameters
$\mathset{\theta_i}$, the Fisher information matrix has components
\begin{equation}
  F_{ij} \equiv \left\langle \frac{\partial \log p\left( \theta | d
    \right)}{\partial \theta_i} \frac{\partial \log p\left( \theta | d
    \right)}{\partial \theta_j} \right\rangle,
\end{equation}
where the average is taken over the data distribution at fixed
$\theta$, $p(d | \theta)$.  The components of the Fisher information
matrix describe the maximum amount of information about the
corresponding parameters available in a given data set; the inverse of
the Fisher information matrix gives the Cramer-Rao bound on the
covariance matrix of unbiased estimators of $\theta$.  For the
likelihood that enters Eq.~\eqref{eq:rate-shape-posterior}, the Fisher
information matrix is 
\begin{multline}
  \mathbf{F} = \left( R_f + R_b \right) \\ \times \begin{pmatrix}
    \left \langle \left(\frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}}\right)^2 \right
    \rangle & \left \langle \frac{\hat{f} \hat{b}}{\left( R_f \hat{f}
      + R_b \hat{b} \right)^2} \right\rangle \\
    \left \langle \frac{\hat{f} \hat{b}}{\left( R_f \hat{f}
      + R_b \hat{b} \right)^2} \right\rangle & \left \langle
    \left(\frac{\hat{b}}{R_f \hat{f} + R_b \hat{b}}\right)^2 \right 
    \rangle
  \end{pmatrix},
\end{multline}
where the expectation values are taken over the distributions
$\hat{f}$ and $\hat{b}$ (i.e.\ they are expectations for one event
from the combined rate distribution).  If the cross-terms are small,
then the Cramer-Rao bound on the uncertainty of $R_f$ will be given by
\begin{equation}
  \sigma_{R_f} \simeq \frac{1}{\sqrt{R_f + R_b}} \left[ \left\langle \left(
    \frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}} \right)^2 \right\rangle\right]^{-1/2}
\end{equation}
Extending a threshold into regions where the factor 
\begin{equation}
  \left( \frac{\hat{f}}{R_f \hat{f} + R_b \hat{b}} \right)^2
\end{equation}
becomes small---that is, into background-dominated
regions---contributes little to reducing the overall uncertainty in the
foreground rate.  Thus, when the background distribution itself is of no interest and computational costs are high, the threshold does not need to be pushed into background-dominated regions in order to obtain an accurate foreground estimate. This is consistent with the behavior of the specific
example in Figure \ref{fig:threshold-rate-dependence}.

@@

To explain this, we will use the same model as described in the
previous subsection, and we will begin with a very specific example.
Let $\Gamma_f = 1$, meaning that the expected number of actual events
with $x > x_1$ is one.  Then there is a $1/64$ chance ($1-\hat{F}(x_{LE}) \approx
1.6\%$) that the loudest event will have $x_{LE} > 4 x_1$.  Consider
this case, and let us also assume that there are no events (noise or
actual) with $x_1 < x < x_{LE}$.

@@

Now, if the true rate really were $\Gamma_f = 64$, then the expected
number of events with $x > x_1$ would be $64$.  So in this case, the
loudest event estimate ignores the fact that there are $\sim 56 -72$
``missing'' events.  However a Bayesian estimate with $x_\mathrm{th}$
set to $x_1$ incorporates this information quite naturally, and so
(correctly) yields an estimated $\Gamma_f$ of order one.
